fMRI by caveat

### 2023-03-21 paper.qmd
updated with figs

@@

{{< video figs/T2star_axial.mov
title="a high-resolution T2*-weighted anatomical scan"
loop=true
width="400"
height="400"
align="right"
>}}

@@

::: {.column-margin}
![Spin-Echo animation from Wikipedia](figs/_7041819_orig.gif){fig-alt="Spin-Echo animation"}
:::

@@

@@

Decoding from a pattern of BOLD responses across voxels which stimulus is shown or experienced by a participant is a sort of 'mind reading'. There are a lot of papers that perform this type of analysis and there are some awesome findings among them. But it's important to think deeply about what we're doing here. As argued in this nice opinion paper[@10.1016/j.tics.2015.07.005], leaving the inferential ....

@@

### References

>>>>

In the end, our data are a bucket of noise from all manner of different sources and we, as a field, have tried to build a house of cards of assumptions and processing strategies to make sense of it. The inferences we can make in this endeavour are always going to be tenuous but just the idea of being able to peer inside someone's living brain and somehow track the brain's activations that give rise to our cognitive processes is miraculous, as is MR imaging itself.

@@

::: {.column-margin}
![T2* anatomical image](figs/T2star_axial.mov){fig-alt="T2* image"}
:::

@@

::: {.column-margin}
![Spin-Echo animation from Wikipedia](figs/_7041819_orig.mov){fig-alt="Spin-Echo animation"}
:::

@@

{{< video figs/conv_animation.mp4
title="Convolution"
loop=true
>}}

@@

Decoding from a pattern of BOLD responses across voxels which stimulus is shown or experienced by a participant is a sort of 'mind reading'. There are a lot of papers that perform this type of analysis and there are some awesome findings among them. But it's important to think deeply about what we're doing here. As argued in this nice opinion paper[@10.1016/j.tics.2015.07.005], having a principled model of how the brain encodes information and computes on it is much more valuable than leaving this important question unanswered. For example, it's quite telling that the earliest work on fMRI 'decoding' was based on visual orientation processing [@10.1038/nn1444], but that the neural and sampling mechanisms that allow us to perform orientation decoding remain under dispute[@10.1146/annurev-vision-093019-111124].

### Caveat XIII: Volumetric atlases and standard spaces make sense only for the subcortex. {#sec-c13}

We all have unique brains, and although their organization is stable at the lower levels of organization, it becomes idiosyncratic very quickly. Especially using volumetric alignment to an (MNI) atlas for cerebral cortex is going to be woefully inadequate in overlaying the detailed functional architecture across individuals. This type of procedure is much better suited for across-subject registration of the subcortex, which has 1. much clearer anatomical areal boundaries in (especially multi-modal anatomical MR images), and 2. a much more stable anatomical organization due to its phylogenetic age.

### Caveat XIV: Surface-based processing is more valid for the cerebral cortex, but also harder to pull off. {#sec-c14}

Using surface-based alignment to a standard surface is better than volumetric alignment to a standard 'atlas' for the cerebral cortex, because it honors its intrinsic architecture. But: the automated segmentation of anatomical images often goes wrong. Because the image value of large veins and gray matter is very similar in standard T1w anatomical images, this means that automated segmentation has a very hard time telling them apart. This results in major problems, where veins and other non-gray-matter anatomical structures are incorporated into the gray matter "ribbon". It helps to add extra images with different MR contrasts such as T2 or FLAIR, but still: the gold standard is manual segmentation, but this can take multiple hours for a single brain.

### Caveat XV: Averaging across subjects is, in the end, impossible. {#sec-c15}

After correct segmentation, surface-based alignment based on cortical folding is passable. But still leaves a lot of functional organization widely variable across individuals. The most sophisticated methods of inter-subject alignment for the cerebral cortex does use these gray-matter surfaces [@10.1038/nature18933]. These methods do a reasonable job but need multiple sessions of anatomical, diffusion, and functional (task & resting state) fMRI to feed into neural network-based areal designations. Anything less than that shoehorns an individual's data into a standard brain or 'atlas' with too little information to fly on.

@@

### Meta-Caveat: Don't believe the hype -- there's always a caveat. {#sec-c16}

_If it sounds too good to be true, it probably is._

The field has progressed through a series of hypes, from decoding[@10.1038/nn1444], to 'dynamic causal modelling'[@10.1016/s1053-811900202-7], to resting-state fMRI[@10.1038/nn1616], to 'laminar' fMRI[@10.1016/j.neuroimage.2017.01.028]. In every one of these hypes there is a nugget of value, and it's become a hype because we as a field are blind to some hitherto unknown caveat. It's our task to value that nugget without being blinded by it, so that we can beat the hype.

# References
